\bf Visual Inference for a Social Network Model

### 2019-02-26 ASApaper/paper-child.Rnw
looks like i forgot to commit stuff

\section{Background} \label{sec:bg}
% \st{
% \begin{enumerate}
% \item significance testing: what is it and why is it important
% \item goodness-of-fit testing: what is it and why is it important
% \item difficulty of these two for sna, why do vis inf
% \end{enumerate}
% }
When selecting and fitting statistical models, there are typiccally three key assessments of interest: significance tests of parameters, goodness-of-fit tests, and power calculations of the tests. With significance testing, the null hypothesis assumes that the data come from a simple model nested within the model of interest. We conduct significance tests of one or more additional parameters to determine if they significantly explain the variability in the data. For goodness-of-fit tests, we examine one or more models of interest to assess how well they fit the data. Finally, power tests quantify the ability of the hypothesis test to detect the difference between the null and alternative hypothesis.  All three of these aspects of statistical modeling are vital for inference. 
%The more complicated the model, the harder it is to determine fit or to include or exclude a parameter.

The more complex the model, however, the harder it is to assess significance, power, and goodness-of-fit. One particularly complicated family of models are those designed to model network change. A \textit{network} is any set of things, such as people, or computers, that are connected in some way, through social relations or the Internet. In a social network, the people are the \textit{nodes}, or \textit{actors}, and the relationships between the people are \textit{edges}, or \textit{ties}. Models for networks are particularly complex, as 
dependencies inherent to network data make them  difficult to model. %Even more challenging is the situation 
And this difficulty only increases when studying \textit{dynamic networks}, the same set of nodes and their changing relationships observed at many points in time. Trying to determine the mechanism underlying change in a network over time only makes the model more complex. Yet, researchers are often interested in modeling dynamic social networks, such as friendship networks among students or collaboration networks between researchers. % X ST a bit less jarring hopefully. X 
Even some of the simplest network models, however, lack the asymptotics required to perform traditional goodness-of-fit tests \citep{holland81}. In additional, direct maximum likelihood estimation of model parameters is frequently impossible due to the intractability of the models \citep{hummeletal}. 
%\hh{is the previous sentence a quote? I'm not sure that I follow.} %\st{it's a paraphrase... shall I make a quote? Also I definitely botched a word. Hopefully makes more sense now? }
% stop 4pm on 1/28/19
In order to circumvent some of these difficulties, we propose new methods for significance testing of parameters, goodness-of-fit testing, and power calculations %of these 
for a set of social network models: stochastic actor-oriented models for dynamic network data, first introduced by \citet{saompaper}. Specifically, we are using \textit{visual inference} in place of traditional statistical methods for social network models, such as Wald or $t$-tests for significance of parameters and in- and outdegree distribution metrics for determining goodness-of-fit. Visual inference, introduced by \citet{Bujaetal}, will allow us to look at the \textit{entire} dataset simulated from a network model as opposed to a (set of) usually one-dimensional metric(s) derived from the network such as outdegree distribution, or a $p$-value for a parameter in the model. By using visual inference to supplement traditional statistical tests, we gain a new understanding of parameters in SAOMs and fit of SAOMs to data.%In addition, we use visual inference to better understand some parameters in a social network model, by looking at how different values for parameters affect the networks simulated from these models.

@@

\subsection{Stochastic Actor-Oriented Models}\label{subsec:saom}

%To model family that we apply to the Senate data is the Stochastic Actor-Oriented Model (SAOM) family. 
Stochastic Actor-Oriented Models (SAOMs) are a family of models for dynamic network data that incorporate both network structure and node-level information to describe how a network observed on two or more occasions changes over time \citep{saompaper}. The two titular properties of SAOMs, stochasticity and actor-orientation, are crucial to understanding networks as they exist naturally: social networks are ever-changing as relationships decay or grow in seemingly random ways, and the actors in them have characteristics that could affect how they change their ties to other nodes in the network. 
% Talk about SAOMs
These unique properties allow for the fitting of some very complicated models to inherently complex data, so it can be exceedingly difficult to interpret parameters and their corresponding estimates. The sheer amount of available parameters to include in the model combined with the difficulty of interpretation make parameter selection and goodness-of-fit testing burdensome as well.

Broadly, a SAOM takes network structure and node covariate information into account in two ways and models the network changes one-at-a-time as a continuous time Markov chain (CTMC). First, the rate of change between states is dictated by a rate function that describes \textit{how often} changes in the network occur, and secondly, the objective function describes \textit{what} those state changes are. As in most other network models, the variables of interest are the edges of the network, $x_{ij}$, where $x_{ij}$ denotes the edge between nodes $i$ and $j$, where $i,j \in \{1, 2, \dots, n=\text{the number of nodes}\}$. 
$x_{ij}$ is modelled as a binary variable, i.e.
  \begin{equation}\label{eq:edgevars}
  x_{ij} =
  \begin{cases}
                                   1 & \text{if an edge from $i$ to $j$ exists} \\
                                   0 & \text{otherwise}
  \end{cases}
  \end{equation}
Edges are treated as \textit{directed}, i.e. in general $x_{ij} \neq x_{ji}$, and self-referencing edges or loops are not allowed, i.e.  $x_{ii} = 0$ for all $i=1, 2, \dots, n$. 
The network is observed $M \geq 2$ times at time points  $t_1 < t_2 < ... < t_M$, and the entire network at time point $t_m$ is denoted as $x(t_m)$. In sections~\ref{rf} and~\ref{of} we discuss the rate and objective functions of a SOAM in more depth.  Additional details on SOAMs can be found in \citet{saompaper, snijders01, snijders:2010, snijders2010, snijdersetall:2007, snijders:2017},

\subsubsection{Rate Function}\label{subsubsec:rf}

All changes in SAOMs are treated as changes made by the nodes, or \textit{actors}, in the network, i.e.\ each actor, $i$, gets a chance to make a change according to the rate function, typically denoted $\lambda_i$, which dictates when relationships between nodes in the network can change. In general, the rate function can take the network structure e.g. outdegree of node $i$, and the node covariates into account, but we use the simple rate function, which is constant over all nodes in a given time period, so that $\lambda_i \equiv \alpha_m$ $\forall i \in \{1, 2, \dots, n\}$. We denote the rate from $t_{m}$ to $t_{m+1}$ as $\alpha_m$ for $m = 1, \dots, M-1$. Using this notation, the \textit{waiting time} to the next chance for actor $i$ to make a change is exponentially distributed with expected value $\alpha_m^{-1}$. Since the rate is the same for all actors, the waiting time for \textit{any} actor to get the oppurtunity to change its set of ties is exponentially distributed with expected value $(n\alpha_m)^{-1}$.

\subsubsection{Objective Function}\label{subsubsec:of}

After actor $i$ has been given the opportunity to change, it probabilistically chooses one of its current ties, $x_{ij}$, to change. The probability that actor $i$ changes its current tie to actor $j$ is determined by the \textit{objective function} of the model and a random component, $U$, which can be thought of as encompassing any other factors that may be influencing the changes the node makes not accounted for by the parameters in the model. %\hh{XXX can you paraphrase the purpose of U in half a sentence?}
%The random component is assumed to have log-Weibull distribution with location parameter $\mu = 0$ and scale parameter $\sigma = 0$ (see \citet{modelsSnijders}), so the probability density function of $U$ is
%\begin{equation}\label{eq:logweibull}
%g(u) = \exp\{-(u + \exp\{-u\})\}.
%\end{equation}
%The objective function, $f_i$, is the driving force of a SAOM. 
Actor $i$ is aiming to maximize its corresponding objective function $f_i$ given the current state of the network, $x$ and the node-level covariates, $\mathbf{Z}$, given as:% which has the form
\begin{equation}\label{eq:objective}
f_i(x, \boldsymbol{\beta}, \mathbf{Z}) = \sum_{k = 1}^K \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}
where $\boldsymbol{\beta} = (\beta_1, \dots, \beta_K)$ are additional model parameters, each associated with some network statistics, $s_{ik}(x, \mathbf{Z})$, $s_{ik}(x, \mathbf{Z})$, calculated with respect to actor $i$ at the current network state $x$. Network statistics range from the simple outdegree, $s_i(x) = \sum_{i\neq j} x_{ij}$, to the more complicated {\it transitive triplets jumping to different covariate}, $s_i(x, \mathbf{Z}) = \sum_{i \neq j \neq h} x_{ij}x_{ih}x_{hj} \cdot \mathbb{I}(z_i = z_h \neq z_j)$. %, plus many more. At last count, in the software we use to fit these models to network data, 
Version 1.2-3 of \texttt{RSiena} \citep{RSiena}, the software used to fit the models here, provides  over 80 possible effects that can be included in the objective function. We discuss these statistics in more detail in Section~\ref{sec:models}.

Objective function $f_{i}(x, \boldsymbol{\beta}, \mathbf{Z})$ and  random component $U$ combine to form the \textit{transition probability}, $p_{ij}$, of the network changing from its current state $x$ to the state with changed tie $x_{ij}$, denoted as $x(i \leadsto j)$:
\begin{equation}\label{eq:transprob}
p_{ij} = \dfrac{\exp\{f_i(x(i \leadsto j), \boldsymbol{\beta}, \mathbf{Z})\}}{\sum_{h} \exp\{f_i(x(i \leadsto h), \boldsymbol{\beta}, \mathbf{Z})\}}
\end{equation}

This probability dictates which edge change is made by the acting node. The acting node can also choose to \textit{not} change at all. This is most likely to occur when the numerator, as calculated for the current state of the network, is larger than for any changes $x(i \leadsto j)$ that could be made.

@@

According to \citet{RSienamanual}, at least two parameters must be included in the objective function: the density and the reciprocity. We denote the density, or out-degree, parameter by $\beta_1$ and the associated statistic as $s_{i1}(x) = \sum_{j} x_{ij}$. Similarly, we denote the reciprocity parameter by $\beta_2$ and the associated statistic as $s_{i2}(x) = \sum_{j} x_{ij}x_{ji}$. We will refer to the very simple model with only these two parameters in the objective function as M1.

@@

\section{Example Data and Models}\label{sec:senatedata}

>>>>

\section{Background} \label{sec:bg}
% \st{
% \begin{enumerate}
% \item significance testing: what is it and why is it important
% \item goodness-of-fit testing: what is it and why is it important
% \item difficulty of these two for sna, why do vis inf
% \end{enumerate}
% }
When selecting and fitting statistical models, there are typically three key assessments of interest: significance tests of parameters, goodness-of-fit tests, and power calculations of the tests. With significance testing, the null hypothesis assumes that the data come from a simple model nested within the model of interest. We then conduct significance tests of one or more additional parameters to determine if they significantly explain the variability in the data. For goodness-of-fit tests, we examine one or more models of interest to assess how well they fit the data. Finally, power tests quantify the ability of the hypothesis test to detect the difference between the null and alternative hypothesis.  All three of these aspects of statistical modeling are vital for inference. 
%The more complicated the model, the harder it is to determine fit or to include or exclude a parameter.

The more complex the model, however, the harder it is to assess significance, power, and goodness-of-fit. One particularly complicated family of models are those designed to model network change. A \textit{network} is any set of things, such as people, or computers, that are connected in some way, through social relations or the Internet. In a social network, the people are the \textit{nodes}, or \textit{actors}, and the relationships between the people are \textit{edges}, or \textit{ties}. Models for networks are particularly complex, as 
dependencies inherent to network data make them  difficult to model. %Even more challenging is the situation 
This difficulty only increases when studying \textit{dynamic networks}, the same set of nodes and their changing relationships observed at many points in time, because of the added temporal dependence. Yet, researchers are often interested in modeling dynamic social networks, such as friendship networks among students or collaboration networks between researchers. % X ST a bit less jarring hopefully. X 
Even some of the simplest network models, however, lack the asymptotics required to perform traditional goodness-of-fit tests \citep{holland81}. In additional, direct maximum likelihood estimation of model parameters is frequently impossible due to the intractability of the models \citep{hummeletal}. 
%\hh{is the previous sentence a quote? I'm not sure that I follow.} %\st{it's a paraphrase... shall I make a quote? Also I definitely botched a word. Hopefully makes more sense now? }
% stop 4pm on 1/28/19
In order to circumvent some of these difficulties, we propose new methods for significance testing of parameters, goodness-of-fit testing, and power calculations %of these 
for one family social network models: continuous time Markov-chain (CTMC) models for dynamic network data, as defined in \citet{saompaper}. Specifically, we are using \textit{visual inference} in place of traditional statistical methods for social network models, such as Wald or $t$-tests for significance of parameters and in- and outdegree distribution metrics for determining goodness-of-fit. Visual inference (VI), introduced by \citet{Bujaetal}, will allow us to look at the \textit{entire} dataset simulated from a network model as opposed to a (set of) usually one-dimensional metric(s) derived from the network, or a $p$-value for a parameter in the model. By using visual inference to supplement traditional statistical tests, we gain a new understanding of the parameters in these CTMC models and ability to assess the fit of the CTMCs to dynamic network data.%In addition, we use visual inference to better understand some parameters in a social network model, by looking at how different values for parameters affect the networks simulated from these models.

@@

\subsection{CTMC Models for Dynamic Social Network Data}\label{subsec:saom}

%To model family that we apply to the Senate data is the Stochastic Actor-Oriented Model (SAOM) family. 
CTMCs are a family of models for dynamic network data that incorporate both network structure and node-level information to describe how a network observed on two or more occasions changes over time \citep{saompaper}. The CTMCs model two key properties: social networks are ever-changing as relationships decay or grow in seemingly random ways, and the actors in them have characteristics that could affect how they change their ties to other nodes in the network. 
% Talk about SAOMs
These unique properties allow for the fitting of some very complicated models to inherently complex data, so it can be exceedingly difficult to interpret parameters and their corresponding estimates. The large amount of available parameters to include in the model combined with the difficulty of interpretation make parameter selection and goodness-of-fit testing burdensome as well.

Broadly, a CTMC model takes network structure and node covariate information into account in two ways and models the network changes one-at-a-time. First, the rate of change between states is dictated by a rate function that describes \textit{how often} changes in the network occur, and second, the objective function describes \textit{what} those state changes are. As in most other network models, the variables of interest are the edges of the network, $x_{ij}$, where $x_{ij}$ denotes the edge between nodes $i$ and $j$, where $i \neq j \in \{1, 2, \dots, n=\text{the number of nodes}\}$. 
$x_{ij}$ is binary, i.e.
  \begin{equation}\label{eq:edgevars}
  x_{ij} =
  \begin{cases}
                                   1 & \text{if an edge from $i$ to $j$ exists} \\
                                   0 & \text{otherwise}
  \end{cases}
  \end{equation}
Edges are treated as \textit{directed}, so that in general $x_{ij} \neq x_{ji}$, and self-referencing edges or loops are not allowed, i.e.  $x_{ii} = 0$ for all $i=1, 2, \dots, n$. 
The network is observed $M \geq 2$ times at time points  $t_1 < ... < t_M$, and the entire network at time point $t_m$ is denoted as $x(t_m)$ for $m \in \{1, \dots, M\}$. We discuss the rate and objective functions of a CTMC model in more depth below.  Additional details on these models can be found in \citet{saompaper, snijders01, snijders:2010, snijders2010, snijdersetall:2007, snijders:2017},

\textbf{Rate Function:} All changes in SAOMs are treated as changes made by the nodes, or \textit{actors}, in the network. Each actor, $i$, gets a chance to make a change according to the rate function, which dictates when relationships between nodes in the network can change. In general, the rate function can take the network structure such as the outdegree of a node, and the node covariates into account, but we use a simple rate function, which is constant over all nodes in a given time period, $\alpha_m$ $\forall i \in \{1, 2, \dots, n\}$. We denote the rate from $t_{m}$ to $t_{m+1}$ as $\alpha_m$ for $m = 1, \dots, M-1$. Using this notation, the \textit{waiting time} to the next chance for actor $i$ to make a change is exponentially distributed with expected value $\alpha_m^{-1}$. Since the rate is the same for all actors, the waiting time for \textit{any} actor to get the oppurtunity to change its set of ties is exponentially distributed with expected value $(n\alpha_m)^{-1}$.

\textbf{Objective Function:} After actor $i$ has been given the opportunity to change, it randomly selects one of its current ties, $x_{ij}$, to change. The probability that actor $i$ changes its current tie to actor $j$ is determined by the \textit{objective function} of the model and a random error term $U$ with log-Weibull distribution \citep{modelsSnijders}. .  %\hh{XXX can you paraphrase the purpose of U in half a sentence?}
%The random component is assumed to have log-Weibull distribution with location parameter $\mu = 0$ and scale parameter $\sigma = 0$ (see \citet{modelsSnijders}), so the probability density function of $U$ is
%\begin{equation}\label{eq:logweibull}
%g(u) = \exp\{-(u + \exp\{-u\})\}.
%\end{equation}
%The objective function, $f_i$, is the driving force of a SAOM. 
Actor $i$ aims to maximize its corresponding objective function $f_i$ given the current state of the network, $x$ and the node-level covariates, $\mathbf{Z}$. This function is defined as:% which has the form
\begin{equation}\label{eq:objective}
f_i(x, \boldsymbol{\beta}, \mathbf{Z}) = \sum_{k = 1}^K \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}
where $\boldsymbol{\beta} = (\beta_1, \dots, \beta_K)$ are additional model parameters, each associated with some network statistics, $s_{ik}(x, \mathbf{Z})$, $s_{ik}(x, \mathbf{Z})$, calculated with respect to actor $i$ at the current network state $x$. Network statistics range from the simple outdegree, $s_i(x) = \sum_{i\neq j} x_{ij}$, to the more complicated {\it transitive triplets jumping to different covariate}, $s_i(x, \mathbf{Z}) = \sum_{i \neq j \neq h} x_{ij}x_{ih}x_{hj} \cdot \mathbb{I}(z_i = z_h \neq z_j)$. %, plus many more. At last count, in the software we use to fit these models to network data, 
Version 1.2-3 of \texttt{RSiena} \citep{RSiena}, the software used to fit the models here, provides over 80 possible effects that can be included in the objective function. We discuss these statistics in more detail in Section~\ref{sec:models}.

Objective function $f_{i}(x, \boldsymbol{\beta}, \mathbf{Z})$ and  random component $U$ combine to form the \textit{transition probability}, $p_{ij}$, of the network changing from its current state $x$ to the state with changed tie $x_{ij}$, denoted as $x(i \leadsto j)$:
\begin{equation}\label{eq:transprob}
p_{ij} = \dfrac{\exp\{f_i(x(i \leadsto j), \boldsymbol{\beta}, \mathbf{Z})\}}{\sum_{h} \exp\{f_i(x(i \leadsto h), \boldsymbol{\beta}, \mathbf{Z})\}}
\end{equation}
This probability dictates which edge change is made by the acting node. The acting node can also choose to \textit{not} change at all. This is most likely to occur when the numerator of Equation~\ref{eq:transprob}, as calculated for the current state of the network, is larger than for any changes $x(i \leadsto j)$ that could be made.

@@

According to \citet{RSienamanual}, at least two parameters must be included in the objective function: the density and the reciprocity. We denote the density, or out-degree, parameter by $\beta_1$ and the associated statistic as $s_{i1}(x) = \sum_{j} x_{ij}$. Similarly, we denote the reciprocity parameter by $\beta_2$ and the associated statistic as $s_{i2}(x) = \sum_{j} x_{ij}x_{ji}$. We will refer to the very simple model with only these two parameters in the objective function as model M1. We define additional models of interest in Section~\ref{sec:models}

@@

\section{Example Data and Models}\label{sec:models}
